---
title: "From Signal Tracks to Replicate-Supported Epigenomic Domain Phenotypes"
author: "Ying ZHANG"
date: "`r Sys.Date()`"
package: epiPortrait
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{From Signal Tracks to Replicate-Supported Epigenomic Domain Phenotypes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(SummarizedExperiment)
library(GenomicRanges)
# rtracklayer BigWig I/O is unsupported on Windows; the core workflow uses the
# shipped tiny BigWigs on Unix/macOS and falls back to example_se on Windows.
use_bw <- .Platform$OS.type != "windows"
```

# Introduction

Epigenomic domains are often summarized by signal enrichment, yet signal
magnitude is only one aspect of domain organization. A focal, high-amplitude
H3K27ac region and a broad, moderately enriched regulatory domain may carry
similar integrated signal while representing different chromatin states.
Likewise, broad H3K4me3 or H3K27me3 domains can expand or contract without
proportional changes in signal abundance. These scenarios represent different
regulatory states but can be obscured when a region is summarized by a single
ranking statistic.

`epiPortrait` was designed to represent these changes explicitly. It separates
**integrated signal magnitude** from **replicate-specific native domain
breadth**, retains within-domain signal dispersion as a complementary
structural descriptor, and aggregates evidence only after each biological
replicate has been evaluated independently.

The resulting domain portrait is **not** intended to replace peak calling or
differential-enrichment statistics. It provides a downstream phenotype layer
for asking whether a domain is predominantly intensity-extreme,
breadth-extreme, jointly extreme, or undergoing a transition between these
states.

## Design principles

The design of `epiPortrait` rests on five principles:

| # | Principle |
|:--|:----------|
| 1 | **Shared coordinates are for comparison, not Breadth truth.** The consensus candidate-domain universe is a comparison frame; it never defines a sample's breadth. |
| 2 | **Breadth comes from replicate-specific native geometry.** Each replicate's own peak/domain calls carry the breadth evidence. |
| 3 | **Intensity and Breadth are separable canonical axes.** They are measured and classified independently, not merged into one rank. |
| 4 | **Replicate evidence is retained before group aggregation.** Per-replicate calls are aggregated with explicit support rules; `Uncertain` is an abstention when evidence is insufficient. |
| 5 | **Biological interpretation depends on the histone mark.** The quantitative engine is mark-independent; the surface terminology is mark-aware. |

## A domain can change in different ways

Across conditions, a domain's architecture state can remodel in distinct
directions:

| Ref | Target | Interpretation |
|:----|:-------|:---------------|
| Typical | Intensity-Super | magnitude-dominant activation |
| Typical | Breadth-Super | breadth expansion |
| Typical | Dual-Super | coupled magnitude + breadth acquisition |
| Intensity-Super | Dual-Super | acquisition of extreme breadth |
| Breadth-Super | Dual-Super | acquisition of extreme magnitude |
| Breadth-Super | Typical | breadth contraction / loss of broad state |

The default `cutoff_scope = "relative"` transition expresses a **relative
rank-state change, not absolute biochemical gain/loss** (see Module 4).

## Conceptual model

`epiPortrait` distinguishes the **shared coordinate frame** (the consensus
candidate-domain universe, one interval per domain across all samples) from the
**sample-specific native peak geometry** (each replicate's own peak/domain
calls). This separation is the central design point: the Breadth-Super call is
derived from the native geometry, never from the consensus width itself.

## Relationship to existing methods

| Tool / family | Primary question | Where epiPortrait differs |
|:--------------|:-----------------|:--------------------------|
| MACS2 / SICER / epic2 | Where are enriched peaks/domains? | epiPortrait starts after domain calling |
| DiffBind / csaw | Where does enrichment differ statistically? | epiPortrait describes domain phenotype and remodeling mode |
| ROSE | Which stitched enhancers are signal-extreme? | epiPortrait separates Intensity from native Breadth and adds replicate support |
| ChIPseeker | Where is a peak relative to genes/features? | epiPortrait uses annotation after quantitative phenotyping |
| deepTools | How do signal tracks look across regions/samples? | epiPortrait converts track measurements into domain-level phenotypes |

`epiPortrait` complements these: it is a replicate-aware quantitative
**domain-phenotyping** layer (Intensity × native Breadth decomposition, plus
within-domain architecture), rather than a peak caller, a differential-caller,
or an annotation engine. Primary use case: histone-mark ChIP-seq / CUT&Tag;
continuous-track workflows also apply to ATAC-seq-derived domains.

Because the input is normalized continuous signal (BigWig) rather than raw
alignments, `epiPortrait` **does not** perform read alignment, duplicate
removal, or count-based differential binding. It is a downstream domain
architecture analysis tool.

# Installation

```{r install, eval = FALSE}
if (!require("BiocManager", quietly = TRUE)) install.packages("BiocManager")
BiocManager::install(c(
  "GenomicRanges", "SummarizedExperiment", "GenomicFeatures",
  "org.Hs.eg.db", "org.Mm.eg.db"
))

devtools::install_github("zying106/epiPortrait", build_vignettes = TRUE)
```

# Quick Start with Example Data

`epiPortrait` ships a simulated `SummarizedExperiment` (`example_se`) with 500
domains across 6 samples (3 Control, 3 Treatment) and 5 quantitative assays:
the two canonical axes (Intensity, native peak breadth geometry), one secondary
architecture descriptor (SignalDispersion), and the two remaining
native-geometry descriptors (NativeMaxPeakWidth, NativeOccupiedWidth,
NativePeakCount).

```{r quickstart}
library(epiPortrait)
data(example_se)
example_se
```

```{r quickstart2}
# Normalization is off by default: input BigWigs should already be
# quantitatively comparable (CPM/RPGC/spike-in). Post-hoc TotalSignal
# rescaling would assume conserved total signal and may remove real global
# biological shifts.
se <- normalize_portrait(example_se, method = "None")

# Main analysis for a multi-condition study: per_group mode, default elbow
# inflection and default log convention.
se <- call_super_domains(se, feature = "Intensity",
                         mode = "per_group", group_var = "Condition",
                         verbose = FALSE)

table(rowData(se)$Intensity_Call__Control)
```

# Module 1: Building the Portrait Matrix

The rest of this vignette runs on a tiny synthetic BigWig dataset shipped in
`inst/extdata` (4 samples: 2 Control, 2 Treatment; 3 domains on chr21), which
exercises the real BigWig I/O and portrait-matrix construction path.
Subsequent inflection-based calling is demonstrated on the 500-domain
`example_se`, because a three-domain I/O fixture is not intended to provide a
statistically meaningful ranked distribution (freeze review 2026-08-11).

## 1.1 The candidate-domain universe is required

Two different "peak files" are involved — do not confuse them.

1. **Candidate-domain universe** (`consensus_peaks`, one shared set for all
   samples) defines *which regions you analyze*. It comes from your peak
   caller's bed/narrowPeak output (MACS2, SICER, epic2, ...), the same
   upstream input ROSE consumes. **Every analysis needs this — including
   single-sample Intensity-only.**
   - *Single sample*: `consensus_peaks` is that sample's own peak file.
   - *Multiple samples*: it is the consensus/union via `get_consensus_peaks()`.
   - Exception: you may import an externally defined domain set (fixed tiling,
     database interval table, or published domains) as a `GRanges` instead.

2. **Per-sample native peaks** (`peak_path`, one file per sample) are each
   sample's *own* peak calls, used to measure native enrichment width per
   sample. **Only the `Breadth` axis needs this**; Intensity-only analyses use
   the candidate universe alone and omit `peak_path`.

**Bottom line: peak files are always required** (to build the candidate
universe) — the only difference is how many and what for. Intensity-only uses
them to define the domains; Breadth additionally uses each sample's own peaks
for width measurement. There is no analysis mode that runs from BigWig alone.

## 1.2 Consensus peaks from your peak-caller output

```{r consensus}
if (use_bw) {
  peak_list <- list(
    rep1 = rtracklayer::import(system.file("extdata", "peaks.bed", package = "epiPortrait")),
    rep2 = rtracklayer::import(system.file("extdata", "peaks.bed", package = "epiPortrait"))
  )
  consensus_peaks <- get_consensus_peaks(peak_list, min_reps = 1)
} else {
  consensus_peaks <- rowRanges(example_se)
}
consensus_peaks
```

## 1.3 Optionally stitch into macro-domains

```{r stitch}
if (use_bw) {
  # For super-enhancer analysis, 12.5 kb is the ROSE-standard stitching distance
  macro_domains <- stitch_epi_peaks(consensus_peaks, stitch_distance = 12500)
} else {
  macro_domains <- consensus_peaks
}
length(macro_domains)
```

## 1.4 Build the portrait matrix

```{r build}
if (use_bw) {
  extdata <- system.file("extdata", package = "epiPortrait")
  samples <- data.frame(
    SampleID  = c("C1", "C2", "T1", "T2"),
    Condition = c("Control", "Control", "Treatment", "Treatment"),
    Replicate = c(1, 2, 1, 2),
    bw_path   = file.path(extdata, c("C1.bw", "C2.bw", "T1.bw", "T2.bw")),
    peak_path = file.path(extdata, c("C1_peaks.bed", "C2_peaks.bed",
                                     "T1_peaks.bed", "T2_peaks.bed"))
  )
  se <- build_portrait_matrix(
    sample_sheet    = samples,
    consensus_peaks = macro_domains,
    workers         = 1
  )
} else {
  se <- example_se
}
se
```

The output `SummarizedExperiment` contains the core assays `Intensity` and
`SignalDispersion`, plus — when per-sample native peak files are provided via
`peak_path` — the sample-specific breadth geometry (`NativeMaxPeakWidth`,
`NativeOccupiedWidth`, `NativePeakCount`). The static genomic interval length
is stored in `rowData(se)$IntervalWidth` (not an assay — it is identical across
samples).

`peak_path` is **optional**. Without it, native breadth geometry and canonical
Breadth-Super calling are unavailable; `Intensity-Super` remains available, and
`SignalDispersion` can still be explored as a secondary architecture
descriptor. When `peak_path` is supplied, the imported native peaks are
retained in `metadata(se)$native_peaks` for auditable peak-level breadth
calling.

For your own data, replace the sample sheet paths with your real normalized
BigWigs (and native peak files), and `consensus_peaks` with your consensus peak
set.

### Custom features

You can extend the portrait with arbitrary per-domain statistics. Example
pattern for real BigWig inputs:

```r
se <- build_portrait_matrix(
  sample_sheet    = samples,
  consensus_peaks = macro_domains,
  custom_features = list(
    Entropy = function(x) {
      p <- (x + 1) / sum(x + 1)
      -sum(p * log2(p))
    }
  )
)
```

# Module 2: Normalization

```{r normalize}
se <- normalize_portrait(example_se, method = "None")
```

| Method | Use case |
|:-------|:---------|
| `None` (default) | Recommended — input BigWigs already quantitatively comparable (CPM/RPGC/spike-in) |
| `TotalSignal` | Rescale to mean domain-set total signal; **assumes conserved total signal and may remove genuine global biological shifts** |
| `Quantile` | Force identical distributions across samples (limma); not recommended for between-condition comparison |
| `TMM` | Count-based scaling (`force_TMM = TRUE` for continuous signal); advanced |

Row-wise Z-score is intentionally **not** an analysis normalization method in
epiPortrait: it would destroy the cross-domain magnitude ranking that Super
calling relies on. Plotting / PCA layers perform display-only scaling
internally.

# Module 3: Super-Domain Calling

`call_super_domains()` ranks domains by a feature and separates super-domains
from typical domains via a hockey-stick inflection point.

```{r calling}
se <- call_super_domains(example_se, feature = "Intensity",
                         method = "elbow", verbose = FALSE)
table(rowData(se)$Intensity_Domain_Type)
```

## 3.1 Inflection methods

Two heuristics are available via the `method` argument:

- **`"elbow"`** (default): the point with the maximum perpendicular distance
  to the first-last ranked line — classic knee detection.
- **`"tangent"`**: a ROSE-inspired tangent-optimization inflection (Whyte et
  al., 2013, *Cell* 153:307-319). Slides a line whose slope is
  `(max - min)/n` and minimizes the number of points below it. Note that this
  is a geometric variant of ROSE's `calculate_cutoff()` step, not a bit-exact
  reproduction; it is only an inflection step, not the full ROSE pipeline.

```{r hockey-methods}
se_elbow <- call_super_domains(example_se, feature = "Intensity",
                               method = "elbow", verbose = FALSE)
se_tangent <- call_super_domains(example_se, feature = "Intensity",
                                 method = "tangent", verbose = FALSE)
table(rowData(se_tangent)$Intensity_Domain_Type)
```

**Explicit top-fraction cutoff (`quantile_cutoff`).** Both axes offer an
explicit top-fraction alternative that bypasses inflection detection: for
`Intensity` it is the given quantile of the ranked domain values; for
`Breadth` it is the same quantile of each replicate's genome-wide eligible
native PeakWidth distribution, so the top `100*(1-quantile_cutoff)`% widest
native peaks are Broad (e.g. `0.95` = top 5% — a common broad-domain
practice). This is an explicit user opt-in with no quality gate (the user
owns the QC choice); it **never** auto-falls-back to, or from, the
data-driven inflection path, so the no-call semantics of the default path are
unchanged.

## 3.2 log-transform is optional

By default `Intensity` and `SignalDispersion` use the package's automatic
`log10(x + 1)` ranking transform (`log_transform = NULL`, auto per feature).
`Breadth` uses the native peak-width calling pipeline described below, and the
static `IntervalWidth` is not auto-log-transformed. Set
`log_transform = FALSE` to rank on the raw scale, or `log_transform = TRUE`
to force it.

**Optional H3K27ac benchmark against ROSE.** If you plan a head-to-head
comparison with the original ROSE method (Whyte et al., 2013, *Cell*
153:307-319), use `method = "tangent"` with `log_transform = FALSE` (raw
signal, ROSE-style scale). Because the tangent implementation is a geometric
variant rather than a bit-exact port, validate any ROSE-reproduction numbers
against the reference `calculate_cutoff()` before drawing quantitative
conclusions. This is an optional benchmark setting, not the package's primary
workflow; the core value of epiPortrait is the Intensity × native-Breadth
decomposition with replicate support. The log-transformed default is retained
as the epiPortrait convention because it stabilises the skewed intensity
distribution for the elbow heuristic and downstream analysis.

## 3.3 Modes

- `mode = "global_consensus"` (default; alias `"consensus"`): calls each
  sample independently, computes replicate support per condition group, and
  requires the support rule to be met in EVERY group (persistent consensus).
- `mode = "per_group"` (recommended for multi-condition studies): each
  replicate is called independently using its own ranked feature distribution;
  replicate-level calls are then aggregated within each condition using the
  selected support rule. One call column per group. The per-domain x
  per-replicate call matrix is available via `get_replicate_calls()`.
  Group-mean ranks are stored for visualization only and never determine the
  group call.
- `mode = "per_sample"`: calls each sample independently.

```{r per-group}
se_pg <- call_super_domains(example_se, feature = "Intensity",
                            mode = "per_group", group_var = "Condition",
                            verbose = FALSE)
grep("Intensity_Call__", colnames(SummarizedExperiment::rowData(se_pg)), value = TRUE)
```

## 3.4 Cutting on breadth: peak-level Breadth-Super

`Breadth` is the peak-level breadth feature. It does **not** rank a
domain x sample assay: instead each replicate's genome-wide eligible native
peak width distribution is cut by an elbow/inflection, broad peaks are mapped
to the shared domains by unique assignment, and the evidence is aggregated
across replicates:

```{r breadth-calling}
se_b <- call_super_domains(example_se, feature = "Breadth",
                           mode = "per_group", group_var = "Condition",
                           verbose = FALSE)
table(rowData(se_b)$Breadth_Call__Control)
```

Key parameters for `feature = "Breadth"`:

- `min_peak_overlap_fraction` (default `0.5`): a native peak is assigned to at
  most one shared domain (the one with maximum overlap) only if at least this
  fraction of its extent lies inside it; peaks below the threshold are
  `Unmapped`, and tied maxima are `Ambiguous` — preventing a broad peak from
  double-counting into neighbouring domains.
- `valid_chroms`: optional vector of allowed chromosomes for the genome-wide
  eligible native peak set.

The peak-level call (PeakWidth / PeakBroadCall / PeakRank) is stored in
`metadata(se)$breadth_peak_calls`, and the unique-assignment provenance in
`metadata(se)$breadth_peak_mapping`.

`IntervalWidth` (alias `Width`) remains available for static region-level
broadness ranking only — it is constant across samples and never defines a
sample-specific breadth.

## 3.5 Combined taxonomy

Combine Intensity and Breadth calls into the four-class taxonomy (plus
`Uncertain` for domains where a call was not reliable):

```{r combine}
se <- call_super_domains(example_se, feature = "Intensity",
                         mode = "per_group", group_var = "Condition",
                         verbose = FALSE)
se <- call_super_domains(se, feature = "Breadth",
                         mode = "per_group", group_var = "Condition", verbose = FALSE)
se <- combine_superdomain_calls(se, group_var = "Condition")
table(rowData(se)$Combined_Class__Control)
```

# Module 4: Condition Transitions

`compare_superdomains()` compares super-domain status between conditions using
`mode = "per_group"` calls (which are replicate-aware within each group):

```{r transitions}
se_t <- call_super_domains(example_se, feature = "Intensity",
                           mode = "per_group", group_var = "Condition",
                           verbose = FALSE)
se_t <- compare_superdomains(se_t, group_var = "Condition",
                             ref_group = "Control", target_group = "Treatment",
                             feature = "Intensity")
table(rowData(se_t)$Intensity_Relative_Transition)
```

Transition classes depend on `cutoff_scope`:

- **`"relative"`** (default): each group is called with its own cutoff. The
  output classes `Relative_Prominence_Up` / `Relative_Prominence_Down` express
  a *relative rank-state* change and **must not** be interpreted as absolute
  signal gain/loss — a domain whose signal is unchanged can still switch class
  if the global background of the treatment shifts the group cutoff.
- **`"reference"`** / **`"pooled"`**: a common cutoff is applied to both
  groups, so `Gain` / `Loss` are closer to absolute interpretations (only
  valid when the BigWigs are quantitatively comparable).

```{r transitions-reference}
se_t2 <- compare_superdomains(se_t, group_var = "Condition",
                              ref_group = "Control", target_group = "Treatment",
                              feature = "Intensity", cutoff_scope = "reference")
table(rowData(se_t2)$Intensity_Transition__reference)
```

`Uncertain` is assigned when a per-group call was not reliable (no-call).

# Module 5: Annotation & Visualization

## 5.1 Domain-aware annotation

```{r annotate}
se <- annotate_epi_domains(se, genome = "hg38")
# rowData: primary_genomic_context, nearest_tss_gene_id,
#          nearest_tss_gene_symbol, nearest_tss_distance_bp,
#          promoter_overlap_gene_count, gene_body_overlap_gene_count,
#          fully_contained_gene_count
head(SummarizedExperiment::rowData(se)$primary_genomic_context)
# long-format links: metadata(se)$domain_gene_links
candidates <- get_domain_genes(se)
head(candidates)
```

## 5.2 Hockey-stick plot

```{r hockey, fig.width = 8, fig.height = 5}
se <- call_super_domains(example_se, feature = "Intensity",
                         method = "tangent", log_transform = FALSE, verbose = FALSE)
plot_hockey_stick(se, feature = "Intensity", label_genes = FALSE)
```

## 5.3 Single-domain feature profile

Quantitative per-domain view across conditions: each feature is faceted with
biological replicates as points and group means as open points; a dashed line
connects the group means so the across-condition direction of change is read
directly. When the object carries combined architecture classes
(`combine_superdomain_calls()`), the domain's class in each condition is
annotated under the axis, linking the raw values to its architectural identity:

```{r portrait, fig.width = 7, fig.height = 5}
plot_domain_feature_profile(example_se, peak_id = rownames(example_se)[1],
                            group_var = "Condition")
```

## 5.4 PCA and correlation QC

```{r qc-plots, fig.width = 8, fig.height = 5}
plot_portrait_pca(example_se, feature = "Intensity", group_var = "Condition")
plot_portrait_correlation(example_se, feature = "Intensity")
```

## 5.5 Academic visualization suite

epiPortrait provides a set of publication-oriented plots that directly
visualise the core method results (Intensity x native-breadth architecture,
replicate support and condition architecture transitions):

```{r academic-viz, fig.width = 8, fig.height = 6}
se_viz <- call_super_domains(example_se, feature = "Intensity",
                             mode = "per_group", group_var = "Condition",
                             verbose = FALSE)
se_viz <- call_super_domains(se_viz, feature = "Breadth",
                             mode = "per_group", group_var = "Condition",
                             verbose = FALSE)
se_viz <- combine_superdomain_calls(se_viz, group_var = "Condition")
se_viz <- compare_superdomain_classes(se_viz,
                                      ref_group = "Control",
                                      target_group = "Treatment")

plot_domain_landscape(se_viz, group = "Control")            # no domain labels by default
plot_class_transition(se_viz, ref_group = "Control", target_group = "Treatment")
plot_replicate_support(se_viz, feature = "Intensity", group = "Control")   # support >= 1/3
plot_domain_class_composition(se_viz)                       # default: super-class focus
plot_domain_class_composition(se_viz, focus = "all")        # full composition incl. Typical

# label the representative domain of each super class (opt-in)
plot_domain_landscape(se_viz, group = "Control", label_n = 5)

# per-group hockey with a descriptive median replicate cutoff
plot_hockey_stick(se_viz, feature = "Intensity", group = "Control",
                  label_genes = FALSE)
```

The landscape is the package signature figure: each domain in the
Intensity x native-breadth plane, coloured by architecture class (pass
`label_n > 0` to name the most representative domain of each super class).
The transition matrix expresses a **relative** architecture-state change
(per-group cutoffs) — it must not be read as absolute gain/loss. Because
non-super domains usually dominate the total, the class-composition plot
defaults to a focused view of the super classes (with their share of all
domains annotated above each bar); pass `focus = "all"` for the full
composition.

## 5.6 Saving and reusing an analysis

The SummarizedExperiment is the scientific record. Save it once, then export
flat tables or re-plot from the object without re-running BigWig I/O:

```{r save-analysis}
f <- tempfile(fileext = ".rds")
saveRDS(se_viz, f)
se2 <- readRDS(f)

# flat tables for Excel / Python / reviewers
outdir <- tempfile("epiPortrait_results_")
dir.create(outdir)
export_epiportrait_results(se2, outdir = outdir)
# a few representative columns (the full table is written to
# epiPortrait_results/domain_results.tsv)
head(get_domain_results(se2, group_var = "Condition")[, c(
  "Domain_ID", "chr", "start", "end", "IntervalWidth",
  "Intensity_Call__Control", "Breadth_Call__Control",
  "Combined_Class__Control", "Combined_Class__Treatment")])
prov <- get_call_provenance(se2, "Intensity")
prov[c("method", "log_transform_used", "cutoff", "call_status")]

unlink(f)
unlink(outdir, recursive = TRUE)
```

## 5.7 Differential domain analysis on continuous signal

Where `compare_superdomains()` reports relative *state* transitions between
discrete phenotypes, `analyze_differential_domains()` adds a statistical
*P-value* for a continuous signal change between condition groups, computed
directly on the integrated BigWig intensity with limma:

```{r differential}
se_diff <- analyze_differential_domains(
  se_viz, feature = "Intensity",
  ref_group = "Control", target_group = "Treatment")
table(rowData(se_diff)$Intensity_DiffStatus)   # Gain / Loss / NS
```

Arbitrary designs (paired, batch-corrected, multi-factor) are supported via
`design` + `contrast`:

```{r differential2, eval=FALSE}
se_diff <- analyze_differential_domains(
  se_diff, feature = "Intensity",
  design  = ~ 0 + Condition + Batch,
  contrast = c(ConditionTreatment = "ConditionTreatment", ...))
```

The volcano view uses the package publication palette and can label the top
domains:

```{r differential-viz}
plot_differential_volcano(se_diff, feature = "Intensity", label_n = 5)
```

**Scope caveat.** The test is on log-transformed integrated continuous signal
(not raw fragment counts) — a defensible approximation for BigWig-native
workflows. It requires >=2 replicates per group (limma residual-variance
estimation). For exact DESeq2/DiffBind-style count statistics, obtain
per-domain counts externally; epiPortrait provides the coordinate universe and
the continuous phenotype layer for that workflow.

# Module 6: Mark-Specific Interpretation

The quantitative engine is mark-independent: every mark is measured on the
same Intensity × native-Breadth axes and classified into the same internal
taxonomy. What changes with the mark is the **biological interpretation** of
those states.

## 6.1 H3K4me3: promoter breadth and transcriptional state

Broad H3K4me3 promoter domains are associated with stable transcriptional
programs, cell-identity genes, and cancer-associated redistribution. With
`epiPortrait`, ask whether a broad promoter is also intensity-extreme, whether
cancer-associated broadening occurs without signal gain, and whether shortening
precedes or accompanies intensity loss:

- `Breadth` captures broad promoter domains;
- `Intensity` captures high-magnitude promoters;
- `Dual-Super` identifies promoters that are both broad and high-magnitude.

`epiPortrait` does **not** claim to discover broad H3K4me3; it decomposes this
known breadth biology into separable quantitative remodeling modes.

## 6.2 H3K27ac: enhancer-domain magnitude versus regulatory territory

A high-ranking enhancer can be strong because of concentrated signal, because
it spans a broad regulatory territory, or because both occur together.
`Intensity-Super`, `Breadth-Super`, and `Dual-Super` express these enhancer
domain phenotypes. For an optional head-to-head comparison with the ROSE
super-enhancer pipeline, use `method = "tangent"` with `log_transform = FALSE`
(a geometric variant, not a bit-exact ROSE reproduction).

## 6.3 H3K27me3 / H3K9me3: broad repressive domains

For broad repressive marks, "Super" is an internal class label; the
mark-appropriate surface terminology is `Intensity-Extreme`,
`Extended-Domain`, and `Dual-Extreme`. Use the upstream broad-domain caller
(e.g. epic2 / SICER) as the native peak source, and keep the preset's
`stitch_distance = 0` sentinel (no re-stitching of the caller-defined
domains).

## 6.4 Application patterns

Organize your analysis around the biological question rather than the mark:

| Pattern | Design | What epiPortrait reports |
|:--------|:-------|:-------------------------|
| Disease-state remodeling | Normal → Cancer | class transitions and width remodeling |
| Drug / perturbation response | Vehicle → Drug, WT → KO | per-condition domain states and transitions |
| Therapy resistance | Sensitive → Resistant | acquisition / loss of super states |
| Differentiation / lineage | State A → State B | state reorganization along the trajectory |
| Cohort recurrence | Independent patients / cell-line models | cross-sample recurrence of a domain phenotype |

For cohort-style datasets, independent patients are **population-level
recurrence evidence, not experimental replicates of a single sample**. Use a
permissive support rule (e.g. `min_replicate_support` tuned to the intended
recurrence threshold, or `support_rule = "all"` when each condition has a
single patient).

## 6.5 Interpretation guardrails

| Claim | Requirement / caveat |
|:------|:----------------------|
| Intensity (between-condition magnitude) | BigWigs must be quantitatively comparable (CPM/RPGC/spike-in) |
| Breadth | Depends on the upstream native peak/domain caller and boundary quality |
| Dual | A conjunction of the two axes, not a new independent measurement |
| SignalDispersion | Describes spatial signal organization, not physical chromatin conformation |
| Annotation | Nearest / overlapping genes are candidates, not causal targets |
| Cohort support | Independent patients are recurrence evidence, not conventional experimental replicates |
| Transition | Default `"relative"` scope is a rank-state change, not absolute biochemical gain/loss |

## 6.6 Input compatibility and parameter guidance

`epiPortrait` expects **normalized** BigWig files that share the same genome
build, bin size, and normalization strategy (e.g., RPGC-, CPM-, or
spike-in-normalized tracks from the same pipeline). Mixing BigWigs from
different sources is risky for cross-sample comparisons; within-sample ranking
and super-domain calling remain interpretable. Positions with **no coverage**
in a BigWig are treated as zero signal during Intensity integration, so domains
with coverage gaps are downward-biased; prefer tracks with complete, consistent
coverage across the analyzed region set and check the
`ZeroFraction` / `NegativeFraction` QC columns from
`check_signal_compatibility()`.

### 6.6.1 Mark-aware preset workflow (e.g. H3K27me3)

`get_mark_preset()` is a configuration helper: it returns mark-appropriate
defaults (for broad repressive marks, `stitch_distance = 0`, i.e. no extra
stitching of the already caller-defined domains, and repressive display
terminology) but the core pipeline is unchanged. A `stitch_distance` of `0`
is a sentinel meaning "do not re-stitch": only call `stitch_epi_peaks()`
when the preset returns a positive distance.

```r
preset <- get_mark_preset("H3K27me3")

# upstream broad-domain caller output imported as GRanges (e.g. epic2/SICER)
domains <- rtracklayer::import("H3K27me3_domains.bed")
# stitch_distance = 0 is a sentinel: keep the upstream caller's geometry
if (preset$stitch_distance > 0) {
  domains <- stitch_epi_peaks(domains, stitch_distance = preset$stitch_distance)
}

se <- build_portrait_matrix(sample_sheet, consensus_peaks = domains)
se <- call_super_domains(se, feature = "Intensity", mode = "per_group")
se <- call_super_domains(se, feature = "Breadth", mode = "per_group")
se <- combine_superdomain_calls(se, group_var = "Condition")

# continuous width remodeling (descriptive, not a significance test)
se <- compute_width_transition(se, ref_group = "Control", target_group = "Treatment")

# repressive display terminology (Extended-Domain / Dual-Extreme)
plot_domain_landscape(se, group = "Control", mark = "H3K27me3")
```

The internal classes remain `Intensity-Super` / `Breadth-Super` /
`Dual-Super`; the `mark` argument only changes surface labels.

## 6.7 Parameter guidance

- Start with `method = "elbow"` (default) for the main analysis; use
  `method = "tangent"` with `log_transform = FALSE` only for an optional
  H3K27ac / ROSE-style benchmark (a geometric variant, not a bit-exact ROSE
  reproduction).
- Use `n_bootstrap` (e.g., 200) for a cutoff **stability interval** (a
  resampling-stability diagnostic, not a classical CI) and success rate.
  For `Intensity` the ranked set is the candidate domains; for `Breadth` it is
  each replicate's eligible native PeakWidth distribution (a per-replicate
  interval is stored in
  `get_call_provenance(se, "Breadth")$replicates[[s]]`). Bootstrap resampling
  is performed on the raw scale with the same transform scale as the main
  analysis. Its information content shrinks with the size of the ranked set;
  at genome-wide scale, check per-replicate cutoff consistency (and, for
  Breadth, peak-set perturbation) instead.
- For multi-condition studies, use `mode = "per_group"` for the main analysis,
  then `compare_superdomains()`. Use `mode = "global_consensus"` (alias
  `"consensus"`) only when the goal is to identify domains reproducibly super
  in **every** condition; consensus support is computed per condition group
  (condition-aware).
- Choose `support_rule` deliberately: `"majority"` (default, needs > n/2
  replicates) is stricter than `"fraction"` 0.5 for n=2 (needs 2/2 vs 1/2).

```{r bootstrap}
se_boot <- call_super_domains(example_se, feature = "Intensity",
                              n_bootstrap = 25, seed = 1)
metadata(se_boot)$superdomain_calls$Intensity$cutoff_stability_interval
```

# Troubleshooting

| Problem | Likely cause | Fix |
|:--------|:-------------|:----|
| `Feature 'X' not found` | Feature name typo | Use `assayNames(se)` / `colnames(rowData(se))` |
| `Breadth requires native peaks` | `feature = "Breadth"` without `peak_path` in the sample sheet | Re-run `build_portrait_matrix()` with per-sample native peak files |
| Negative-signal error | BigWig contains negative values | Provide non-negative (CPM/RPGC/spike-in) tracks or `negative_policy = "clip_zero"` |
| `Constant feature values` | All domains identical for a feature (e.g., `IntervalWidth` in a single-sample set) | Use a dynamic feature like `Intensity` |
| Super count unexpectedly high | log-transformed scale + tangent | Try `log_transform = FALSE` |
| No reliable inflection (`no_call`) | Curve lacks a clear break | **Intensity:** consider an explicit `quantile_cutoff` (top fraction of ranked domain values) or inspect the hockey-stick plot. **Breadth:** `quantile_cutoff` is available as an explicit top-fraction opt-in on the genome-wide native PeakWidth distribution (no quality gate — the user owns the QC choice); the default data-driven path keeps no-call as no-call, and the two paths never auto-fall-back to each other |

# Session Info

```{r session}
sessionInfo()
```
