---
title: "epiPortrait: 4D Geometric Profiling of Epigenetic Landscapes"
author:
- name: "Ying ZHANG"
  affiliation: "Zhejiang University"
  email: "zhangying@example.com"
  orcid: "0009-0005-9644-7062"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
vignette: >
  %\VignetteIndexEntry{epiPortrait: 4D Geometric Profiling of Epigenetic Landscapes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
table {
  border-collapse: collapse;
  width: 100%;
}
th, td {
  border: 1px solid #ddd;
  padding: 8px;
}
th {
  background-color: #f2f2f2;
  text-align: left;
}
.comparison-table-wrap {
  overflow-x: auto;
}
</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width  = 8,
  fig.height = 5,
  message    = FALSE,
  warning    = FALSE
)
```

---

# Introduction

`r BiocStyle::Biocpkg("epiPortrait")` is a computational framework for
analyzing epigenetic signals (ChIP-seq, ATAC-seq, CUT&Tag) beyond their
one-dimensional read count. It extracts a **four-dimensional geometric
portrait** from BigWig coverage profiles — **Width**, **Intensity**, **Height**,
and **Skewness** — at every genomic domain, then applies multivariate
statistical testing (MANOVA) to detect **Shape Shift** events where the
physical conformation of chromatin changes between experimental conditions.

## Key Features & Capabilities

1. **4D Feature Extraction** — Parse BigWig coverage into Width, Intensity,
   Height, and background-gated Skewness for every consensus domain in parallel.

2. **Multivariate Shape Shift Testing** — MANOVA-based global test across all
   four dimensions with partial eta-squared effect sizes. Supports simple
   two-group, paired/blocked, and complex multi-factor designs.

3. **Shape Classification** — Translates abstract Pillai trace scores into six
   biologically interpretable categories (Concentration, Flattening, Polarity
   Shift, Global Gain, Global Loss, Complex).

4. **Conformational Outlier Detection** — Mahalanobis distance with robust
   (MCD) covariance estimation identifies domains with intrinsically unusual
   4D geometry, without requiring group labels.

5. **Temporal & Dose-Response** — Polynomial contrast MANOVA detects monotonic
   broadening, transient sharpening, and non-linear drift across timepoints.

6. **Power Analysis** — Monte Carlo simulation of MANOVA power across sample
   sizes and effect sizes for prospective experimental design.

7. **Publication-Ready Visualization** — 4D volcano plots, radar charts, group
   overlay tracks, hockey stick plots, Manhattan plots, PCA, and correlation
   heatmaps.

8. **Complete Pipeline** — From consensus peak calling to annotation,
   enrichment, BED/IGV export, parameterised HTML report, and automated
   biological narrative synthesis.

## Comparison with Related Approaches

<div class="comparison-table-wrap">

| Feature | epiPortrait | DiffBind | ChIPseeker | csaw | ROSE |
|:--------|:------------|:---------|:-----------|:-----|:-----|
| **Signal dimensions** | 4D (W,I,H,S) | 1D (counts) | 1D (peaks) | 1D (counts) | 1D (intensity) |
| **Multivariate testing** | &#10004; MANOVA | &#10008; | &#10008; | &#10008; | &#10008; |
| **Shape shift detection** | &#10004; | &#10008; | &#10008; | &#10008; | &#10008; |
| **Domain shape classification** | &#10004; 6 classes | &#10008; | &#10008; | &#10008; | &#10008; |
| **Super-enhancer detection** | &#10004; W/I/H | &#10008; | &#10008; | &#10008; | &#10004; Intensity |
| **Temporal trend testing** | &#10004; polynomial | &#10008; | &#10008; | &#10008; | &#10008; |
| **Outlier detection (no groups)** | &#10004; Mahalanobis | &#10008; | &#10008; | &#10008; | &#10008; |
| **Power analysis** | &#10004; | &#10008; | &#10008; | &#10008; | &#10008; |
| **IGV session export** | &#10004; | &#10008; | &#10008; | &#10008; | &#10008; |
| **Narrative synthesis** | &#10004; | &#10008; | &#10008; | &#10008; | &#10008; |

</div>

&#10004; = Supported | Partial = Conditionally supported | &#10008; = Not supported

---

# Installation

```{r install, eval = FALSE}
if (!require("BiocManager", quietly = TRUE)) install.packages("BiocManager")
BiocManager::install(c(
  "GenomicRanges", "SummarizedExperiment", "rtracklayer",
  "ChIPseeker", "GenomicFeatures", "limma",
  "clusterProfiler", "org.Hs.eg.db", "org.Mm.eg.db"
))

# Install epiPortrait from GitHub
devtools::install_github("zhangying/epiPortrait", build_vignettes = TRUE)
```

```{r load}
library(epiPortrait)
library(SummarizedExperiment)
library(ggplot2)
```

---

# Quick Start with Example Data

`epiPortrait` ships with a simulated dataset containing 500 peaks across
6 samples (3 Control, 3 Treatment) with **known ground-truth shape shifts**
embedded in the data. This enables method validation against a known answer.

```{r example-data}
data(example_se)
example_se

# Ground truth: 5 types of shape shift embedded
table(mcols(rowRanges(example_se))$True_Class)
```

---

# Module 1: Building the 4D Portrait Matrix

The core engine `build_portrait_matrix()` reads BigWig coverage profiles and
constructs a `SummarizedExperiment` with 4 assays (plus optional custom
features).

## Full Pipeline

```{r build-matrix, eval = FALSE}
sample_sheet <- data.frame(
  SampleID  = c("Ctrl_1", "Ctrl_2", "Ctrl_3", "Treat_1", "Treat_2", "Treat_3"),
  Condition = c("Control", "Control", "Control", "Treatment", "Treatment", "Treatment"),
  bw_path   = c("ctrl1.bw", "ctrl2.bw", "ctrl3.bw", "treat1.bw", "treat2.bw", "treat3.bw")
)

# Consensus peaks from your peak-caller output
# consensus_peaks <- get_consensus_peaks(peak_list, min_reps = 2)

# Optionally stitch into macro-domains
# macro_domains <- stitch_epi_peaks(consensus_peaks, stitch_distance = 12500)

# Build the 4D portrait
se <- build_portrait_matrix(
  sample_sheet    = sample_sheet,
  consensus_peaks = consensus_peaks,
  workers         = 4,
  bg_quantile     = 0.1,
  on_disk         = FALSE,
  cache_dir       = "epiPortrait_cache"
)

# Normalize
se <- normalize_portrait(se, method = "TotalSignal")
```

```{r example-se, echo = FALSE}
se <- example_se
```

## Deep Dive: Feature Extraction Engine

**Width** is constant across all samples for each peak (the physical span
of the consensus domain). It is stored once and replicated for matrix
compatibility.

**Intensity** is the total signal abundance: `sum(coverage_view)`. This is
analogous to traditional read count but computed from BigWig coverage rather
than count tables, enabling precise quantification within arbitrary
domain boundaries.

**Height** is the maximum signal density at the summit: `max(coverage_view)`.
This captures peak sharpness independent of total signal.

**Skewness** is computed via `moments::skewness()` on coverage values
exceeding the `bg_quantile` threshold within each domain (the **Valley Trap**
method). This gate excludes background noise that would otherwise dominate
the skewness estimate in low-signal regions.

**Custom Features**: You can extend the portrait matrix with arbitrary
user-defined geometric dimensions via the `custom_features` parameter.
Each custom function receives the coverage view for one peak as a
`NumericList` and must return a single numeric value.

```{r custom-features, eval = FALSE}
se <- build_portrait_matrix(
  sample_sheet    = samples,
  consensus_peaks = peaks,
  custom_features = list(
    Entropy = function(x) {
      p <- (x + 1) / sum(x + 1)
      -sum(p * log2(p))
    },
    Coverage_Breadth = function(x) sum(x > quantile(x, 0.5)) / length(x)
  )
)
```

## Caching Strategy

Setting `cache_dir` to a directory path caches per-sample BigWig coverage
views as RDS files keyed by a hash of the BigWig path and consensus peaks.
Subsequent calls with identical inputs skip BigWig I/O entirely, making it
practical to tune `bg_quantile` or iterate on feature engineering without
re-reading multi-gigabyte BigWig files.

---

# Module 2: Normalization

```{r norm}
se <- normalize_portrait(se, method = "TotalSignal")
```

| Method | Description | Use Case |
|:-------|:------------|:---------|
| `TotalSignal` | Scales each sample to equal total signal across all domains | Default; robust to most scenarios |
| `Quantile` | Forces identical empirical distributions across samples | Severe batch effects |
| `Z-score` | Standardizes each domain to mean = 0, sd = 1 per sample | Heatmaps and clustering ONLY; not for differential testing |
| `TMM` | Trimmed mean of M-values (edgeR) | Count data only; requires `force_TMM = TRUE` for BigWig signals |

---

# Module 3: Quality Control

PCA and correlation heatmaps on individual geometric features reveal sample
clustering, batch effects, and outlier samples before formal statistical
testing.

```{r qc-pca, fig.width = 9, fig.height = 5}
plot_portrait_pca(se, feature = "Intensity", group_var = "Condition", n_top = 500)
```

```{r qc-corr, fig.width = 7, fig.height = 6}
plot_portrait_correlation(se, feature = "Intensity")
```

---

# Module 4: Statistical Testing

## 4a. Global Shape Shift Test (MANOVA)

`test_global_shape_shift()` performs a per-peak MANOVA across all four
geometric dimensions (Width, Intensity, Height, Skewness) simultaneously.
This is the **primary innovation** of epiPortrait: detecting multivariate
conformational change that is invisible to any single-dimension test.

### Simple Two-Group Comparison

```{r shift-simple}
shift_res <- test_global_shape_shift(se, group_var = "Condition", workers = 1)
head(shift_res[, c("Peak_ID", "Shape_Shift_Score", "partial_eta_sq", "adj.P.Val")], 5)
```

The `partial_eta_sq` column reports the proportion of multivariate variance
explained by group membership — a Cohen-style effect size directly
comparable across peaks.

### Complex Multi-Factor Designs

```{r shift-formula}
shift_res <- test_global_shape_shift(
  se,
  formula = ~ Condition,
  term    = "Condition",
  workers = 1
)
```

The `formula` interface accepts any model specification from `colData(se)`:
`~ genotype * treatment + batch`, `~ timepoint + patient`, etc. The last
term in the formula is auto-selected as the effect of interest unless
overridden by `term`.

## 4b. Per-Dimension Differential Testing

```{r diff}
diff_res <- test_differential_feature(
  se,
  feature      = "Intensity",
  group_var    = "Condition",
  target_group = "Treatment",
  ref_group    = "Control",
  method       = "limma"
)

# Test Height similarly
diff_height <- test_differential_feature(
  se, feature = "Height",
  group_var = "Condition",
  target_group = "Treatment", ref_group = "Control",
  method = "limma"
)
```

**Method selection guide:**

| Method | Pros | Cons | Recommended for |
|:-------|:-----|:-----|:----------------|
| `limma` | Empirical Bayes moderation; stable with n < 5 | Assumes normality | Default; most scenarios |
| `t.test` | No prior distribution assumptions | Underpowered with small n | n ≥ 6 per group |
| `wilcox` | Distribution-free; robust to outliers | Low power with n < 8 | Non-normal data with sufficient replicates |

---

# Module 5: Shape Classification

```{r classify}
shift_res <- classify_shape_shift(
  se, shift_res,
  group_var    = "Condition",
  target_group = "Treatment",
  ref_group    = "Control",
  fc_threshold     = 0.5,
  skewness_threshold = 0.3
)

table(shift_res$Shape_Class)
```

### Classification Rules

Each peak is classified by comparing per-dimension log2 fold changes and
skewness delta against user-defined thresholds:

| Class | Intensity FC | Height FC | Skewness Delta | Interpretation |
|:------|:------------|:----------|:---------------|:---------------|
| **Concentration** | Stable | Up (> threshold) | Stable | Signal concentrates at summit |
| **Flattening** | Stable | Down (< -threshold) | Stable | Signal spreads laterally |
| **Polarity Shift** | Stable | Stable | Large (> threshold) | Asymmetric redistribution |
| **Global Gain** | Up | Concordant or stable | Stable | Coordinated activation |
| **Global Loss** | Down | Concordant or stable | Stable | Coordinated repression |
| **Complex** | ≥ 2 dimensions discordant | | | Multi-dimensional remodelling |
| **Stable** | All stable | All stable | Stable | No significant change |

---

# Module 6: Temporal & Dose-Response Analysis

```{r temporal, eval = FALSE}
# Requires ≥ 3 ordered timepoints
se$Timepoint <- c(0, 0, 2, 2, 6, 6, 24, 24)

temporal_res <- test_temporal_shape_shift(
  se,
  time_var   = "Timepoint",
  block_var  = "Patient",    # optional, for paired designs
  poly_order = 2,            # linear + quadratic trends
  workers    = 4
)

# Output columns per trend: Linear_Score, Linear_Pval, Linear_Eta2,
#                           Quadratic_Score, Quadratic_Pval, Quadratic_Eta2
head(temporal_res[, grep("Pval|Peak_ID", colnames(temporal_res))])
```

**Interpretation:**

- **Linear trend**: Monotonic broadening or sharpening across time/dose.
- **Quadratic trend**: Peak-then-collapse (or dip-then-rise) — non-linear
  temporal dynamics often associated with transient regulatory events.
- **Cubic trend** (poly_order = 3): Oscillatory patterns.

---

# Module 7: Conformational Outlier Detection

`detect_conformational_outliers()` identifies domains whose 4D geometry
deviates from the population distribution using the Mahalanobis distance.
Unlike MANOVA, this does **not** require group labels — it detects
intrinsically unusual conformations, making it suitable for single-sample
analysis and QC.

```{r outlier}
outliers <- detect_conformational_outliers(se, method = "robust")

head(outliers[, c("Peak_ID", "Mahalanobis_D2", "P.Value", "adj.P.Val", "Outlier_Flag")], 5)
```

**Robust vs Classic estimation:**

| Estimator | Method | Outlier Resistance | Use Case |
|:----------|:-------|:-------------------|:---------|
| `"robust"` | MCD (Minimum Covariance Determinant) | High — outliers excluded from covariance estimation | Default; most scenarios |
| `"classic"` | Sample mean and covariance (MLE) | None — outliers inflate covariance | Only when MCD fails (e.g., n < 8) |

**Single-sample mode:** Provide `ref_samples` to estimate the reference
distribution from a control cohort and evaluate a query sample against it.

```{r outlier-single, eval = FALSE}
outliers <- detect_conformational_outliers(
  se,
  method      = "robust",
  ref_samples = c("Normal_1", "Normal_2", "Normal_3")
)
```

---

# Module 8: Visualization Suite

## 8a. 4D Volcano Plot {.tabset}

X-axis: Intensity log2 fold-change | Y-axis: -log10(adj.P.Val) |
Point size: domain width | Point color: skewness shift.

```{r volcano, fig.width = 10, fig.height = 8}
plot_portrait_volcano(
  diff_res, se,
  target_group = "Treatment",
  ref_group    = "Control",
  p_cutoff     = 0.05,
  fc_cutoff    = 1
)
```

**Plot key:**

| Aesthetic | Data | Interpretation |
|:----------|:-----|:---------------|
| X position | Intensity log2FC | Traditional differential binding signal |
| Y position | -log10(adj.P.Val) | Statistical significance |
| Point size | log10(Width + 1) | Physical domain span |
| Point color | Skewness_Delta | Asymmetric redistribution (red = right-skewed in target, blue = left-skewed) |
| Point alpha | Significance category (Up/Down/Not Sig) | Non-significant points fade to background |

## 8b. Single-Peak Radar Portrait

Visualize the 4D conformation of one peak across experimental groups using
Min-Max scaled polar coordinates. All four dimensions are normalized to
[0, 1] for visual comparability — the **shape**, not absolute magnitude,
is emphasized.

```{r radar, fig.width = 7, fig.height = 6}
top_peaks <- shift_res$Peak_ID[shift_res$adj.P.Val < 0.05]
if (length(top_peaks) > 0) {
  plot_peak_portrait(se, peak_id = top_peaks[1], group_var = "Condition")
}
```

## 8c. Shape Shift Validation: Group Overlay Tracks

The most intuitive validation: mean coverage of two groups overlaid on the
same panel with ± SEM ribbons.

```{r overlay, fig.width = 10, fig.height = 5, warning = FALSE, eval = FALSE}
# Requires actual BigWig files on disk (example_se uses simulated paths).
# With your own data, this produces the most intuitive validation figure.
if (length(top_peaks) > 0) {
  plot_shift_comparison(
    se, peak_id = top_peaks[1],
    group_var = "Condition",
    group1    = "Control",
    group2    = "Treatment",
    extend_bp = 1000
  )
}
```

## 8d. Hockey Stick Plot (Super-Domain Identification)

ROSE-style ranking: domains are ordered by a geometric feature, and the
inflection point separates **Super-Domains** from **Typical** domains.

```{r hockey, fig.width = 8, fig.height = 5}
se <- call_super_domains(se, feature = "Intensity")
plot_hockey_stick(se, feature = "Intensity", top_n_label = 10)

# For broad-domains based on width:
# se <- call_super_domains(se, feature = "Width")
# plot_hockey_stick(se, feature = "Width")

# For steep-peaks based on height:
# se <- call_super_domains(se, feature = "Height")
# plot_hockey_stick(se, feature = "Height")
```

## 8e. Manhattan Plot of Conformational Outliers

Genome-wide view of conformational outlier significance, with chromosome
alternation and significance threshold.

```{r manhattan, fig.width = 12, fig.height = 5}
outliers <- detect_conformational_outliers(se, method = "robust")
plot_outlier_manhattan(outliers, y_axis = "logP", label_n = 10)
```

**Available visualizations by analysis stage:**

| Stage | Function | Description |
|:------|:---------|:------------|
| QC | `plot_portrait_pca()` | PCA on any geometric dimension |
| QC | `plot_portrait_correlation()` | Sample Pearson correlation heatmap |
| Differential | `plot_portrait_volcano()` | 4D volcano: logFC × P × Width × Skewness |
| Differential | `plot_peak_portrait()` | Radar chart of single-peak 4D conformation |
| Validation | `plot_shift_comparison()` | Group overlay: mean ± SEM tracks |
| Validation | `plot_peak_track()` | Per-sample faceted coverage profiles |
| Calling | `plot_hockey_stick()` | ROSE-style super-domain ranking |
| Outlier | `plot_outlier_manhattan()` | Genomic distribution of conformational outliers |
| Power | `plot_power_curve()` | Power × sample size × effect size |

---

# Module 9: Annotation & Enrichment

## 9a. Genomic Annotation

Map shape-shifted domains to nearest genes using ChIPseeker's annotation
engine with smart lazy-loading for common genomes.

```{r annotation, eval = FALSE}
# Data frame output (merges annotations with epiPortrait feature columns)
anno_df <- annotate_epi_peaks(
  shift_res[shift_res$adj.P.Val < 0.05, ],
  genome     = "hg38",
  tss_region = c(-3000, 3000)
)

# Raw csAnno object for ChIPseeker plotting
anno_obj <- annotate_epi_peaks(
  sig_peaks, genome = "hg38",
  return_format = "csAnno"
)
ChIPseeker::plotAnnoPie(anno_obj)
```

**Supported genomes:** `"hg38"`, `"hg19"`, `"mm10"`, or `"custom"` (with
user-supplied `custom_txdb` and `custom_annoDb`).

## 9b. Functional Enrichment

Run GO (BP, MF, CC) and KEGG pathway enrichment **per Shape_Class category**,
enabling class-specific functional interpretation.

```{r enrichment, eval = FALSE}
enrich_res <- enrich_shape_shifted(shift_res, organism = "human")

# Access class-specific results
enrich_res$go$Concentration  # GO terms enriched in Concentration peaks
enrich_res$go$Flattening     # GO terms enriched in Flattening peaks
```

## 9c. Automated Biological Narrative

`summarize_findings()` synthesizes classified shape-shift results into a
structured, human-readable narrative suitable for a manuscript results
section — top-level summary, per-class breakdown with gene examples, top
peaks table, enrichment interpretation, and recommended follow-up experiments.

```{r narrative, eval = FALSE}
narrative <- summarize_findings(
  shift_res, se,
  enrichment_res = enrich_res,
  fdr_cutoff     = 0.05
)

# Copy-paste ready full text
cat(narrative$full_text)

# Or extract individual sections
cat(narrative$summary)           # one-paragraph top-level summary
cat(narrative$class_breakdown)   # per-class narrative with gene examples
cat(narrative$follow_up)         # recommended next experiments
```

---

# Module 10: Power Analysis

Before designing your experiment, determine the minimum number of biological
replicates needed to detect a shape shift of a given magnitude.

```{r power, eval = FALSE}
power_df <- simulate_power(
  n_range    = 3:10,
  eta2_range = c(0.06, 0.14, 0.30, 0.50),
  n_dims     = 2,
  n_sim      = 200,
  workers    = 4
)

plot_power_curve(power_df)
```

**Interpreting partial eta² in the multivariate context:**

| eta² | Benchmark | Description |
|:-----|:----------|:------------|
| 0.01 | Small | Subtle shape shift; requires n > 20 per group |
| 0.06 | Medium | Detectable with n = 8–12 per group |
| 0.14 | Large | Detectable with n = 4–6 per group |
| 0.30+ | Very large | Detectable with n = 3 per group |

---

# Module 11: Export

## BED File Export

```{r bed-export, eval = FALSE}
export_shifted_bed(
  shift_res,
  file       = "shape_shifted_regions.bed",
  fdr_cutoff = 0.05
)
```

The output BED file includes the `Shape_Class` column for downstream
filtering and coloring in IGV.

## IGV Session Export

Double-click the generated XML to open all BigWig tracks and a
shape-class-colored annotation track in IGV.

```{r igv, eval = FALSE}
export_igv_session(
  shift_res, se,
  genome       = "hg38",
  session_file = "epiPortrait_session.xml"
)
```

This creates:
- `epiPortrait_session.xml` — IGV session with BigWig tracks + annotation track
- `epiPortrait_session_annotation.bed` — shape-shifted regions colored by `Shape_Class`

---

# Module 12: One-Click Parameterised Report

Generate a complete, publication-ready HTML report with all analysis results,
figures, and tables in a single call. Modules auto-hide when their parameters
are left at `NULL`.

```{r report, eval = FALSE}
render_report(
  sample_sheet          = sample_sheet,
  consensus_peaks       = consensus_peaks,
  genome                = "hg38",
  target_group          = "Treatment",
  ref_group             = "Control",
  super_domain_feature  = "Intensity",
  output_file           = "epiPortrait_analysis.html"
)
```

You can also create an R Markdown document from the built-in template:
*File > New File > R Markdown... > From Template > epiPortrait Report*.

---

# Application Scenarios {#scenarios}

This section illustrates how epiPortrait's modules combine to address
specific biological questions across common experimental designs.

## Scenario A: Drug Treatment Reshapes the H3K27ac Regulatory Landscape {.tabset}

**Biological context.** A compound inhibits a chromatin remodeller. Standard
differential binding (DiffBind/csaw) on H3K27ac ChIP-seq identifies 1,200
gained and 800 lost peaks at FDR < 0.05. But does the treatment also change
the **physical architecture** of regulatory elements — making enhancers
broader, sharper, or asymmetrically distributed?

### Analysis Strategy

1. **Portrait matrix** on consensus peaks from all samples
2. **MANOVA shape shift test** across all four dimensions — this detects
   conformational changes missed by DiffBind
3. **Classify** shifted regions into concentration, flattening, polarity
   shift, gain, loss, and complex
4. **Overlay tracks** validate the top candidates per class
5. **Enrichment** per shape class reveals distinct biology: flattening in
   super-enhancers of identity genes, concentration at signalling pathway
   promoters

```{r scenario-a, eval = FALSE}
# After building the portrait matrix and running MANOVA + classification:
shift_res <- classify_shape_shift(se, shift_res,
  target_group = "Drug", ref_group = "DMSO")

# Compare: how many shape-shifted peaks are NOT differentially bound?
diff_peaks <- rownames(diff_res)[diff_res$adj.P.Val < 0.05]
shape_peaks <- shift_res$Peak_ID[shift_res$adj.P.Val < 0.05]
shape_only <- setdiff(shape_peaks, diff_peaks)

cat(sprintf(
  "%d peaks with shape shift but no significant change in total Intensity.\n",
  length(shape_only)
))

# Flattening events at super-enhancers
flattened <- shift_res[shift_res$Shape_Class == "Flattening" & shift_res$adj.P.Val < 0.05, ]
# These are candidate regions where the drug causes regulatory broadening
```

### Interpretation Checklist

| Question | epiPortrait Answer |
|:---------|:-------------------|
| Are there shape changes invisible to DiffBind? | Compare `shape_peaks` vs `diff_peaks` overlap |
| Which shape class dominates the treatment response? | `table(shift_res$Shape_Class)` on significant hits |
| Are flattened regions enriched at super-enhancers? | Cross-reference with `call_super_domains(se, "Intensity")` |
| Do concentration events map to acute-response genes? | `enrich_shape_shifted()` GO enrichment per class |

## Scenario B: Disease vs Normal — Finding Latent Regulatory Remodelling {.tabset}

**Biological context.** ATAC-seq on tumour vs adjacent normal tissue. Most
analyses focus on differentially accessible peaks. However, tumour
epigenomes often exhibit **broadening of accessibility domains** at
oncogene loci and **sharpening at tumour-suppressor promoters** —
conformational changes that may not produce large logFC in total
accessibility.

### Analysis Strategy

1. Build the portrait matrix on a **comprehensive consensus peak set**
   (union of all peaks, filtered for reproducibility ≥ 2 samples)
2. Avoid stitching — tumour-specific peaks may be biologically real, and
   stitching can merge them with background
3. MANOVA with `block_var = "Patient"` if paired tumour-normal samples exist
4. Focus classification on `Flattening` and `Concentration` — these are the
   most interpretable in cancer epigenetics
5. Overlay tracks for the top 5 per class
6. Export IGV session for oncologist review

```{r scenario-b, eval = FALSE}
# Paired tumour-normal design
shift_res <- test_global_shape_shift(
  se,
  group_var = "Condition",
  block_var = "Patient",
  test      = "Pillai"
)

# Classification with stricter thresholds for clinical relevance
shift_res <- classify_shape_shift(
  se, shift_res,
  target_group = "Tumour",
  ref_group    = "Normal",
  fc_threshold = 0.7,          # more stringent
  skewness_threshold = 0.4
)

# Prioritise: Flattening at known oncogene loci
flattened <- shift_res[
  shift_res$Shape_Class == "Flattening" &
  shift_res$adj.P.Val < 0.01,
]
```

### Key Considerations for Clinical Cohorts

- **Batch effects**: If samples were processed in multiple batches, include
  `batch` as a covariate: `formula = ~ batch + Condition`
- **Sample size**: For n < 5 per group, prefer the `Pillai` test statistic
  (most robust to unequal covariance) and `partial_eta_sq` as the primary
  ranking metric rather than P-value alone
- **Replication**: Shape-shifted peaks at known disease loci should be
  validated by ChIP-qPCR of the summit region, specifically measuring
  peak shape (not just enrichment)

## Scenario C: Time-Course Differentiation — Detecting Temporal Shape Drift {.tabset}

**Biological context.** ATAC-seq or H3K27ac at 0, 2, 6, 24, and 72 hours
of differentiation. The standard approach (pairwise DE at each timepoint
vs 0h) generates 4 separate lists with no model of temporal coherence.
`test_temporal_shape_shift()` uses polynomial contrasts to partition
multivariate variance into linear, quadratic, and cubic trend components.

### Analysis Strategy

1. Build portrait matrix across all timepoints
2. `test_temporal_shape_shift()` with `poly_order = 3`
3. Classify peaks by their dominant trend type:
   - **Linear**: monotonic broadening (enhancer maturation) or sharpening
     (promoter commitment)
   - **Quadratic**: transient opening at pioneer factor targets, peaking
     at intermediate timepoints
   - **Cubic**: oscillatory remodelling — often at cell-cycle or circadian
     loci
4. Per-trend functional enrichment

```{r scenario-c, eval = FALSE}
temporal_res <- test_temporal_shape_shift(
  se,
  time_var   = "Hour",
  poly_order = 3,
  workers    = 4
)

# Categorise by dominant trend
temporal_res$Dominant_Trend <- apply(temporal_res, 1, function(row) {
  pvals <- c(
    Linear    = as.numeric(row["Linear_Pval"]),
    Quadratic = as.numeric(row["Quadratic_Pval"]),
    Cubic     = as.numeric(row["Cubic_Pval"])
  )
  if (all(is.na(pvals))) return("None")
  names(which.min(pvals))
})

table(temporal_res$Dominant_Trend)

# Quadratic trend: transient opening — candidate pioneer factor targets
quad_peaks <- temporal_res[
  temporal_res$Dominant_Trend == "Quadratic" &
  temporal_res$adj.P.Val < 0.05,
]
```

### Interpretation of Polynomial Trends

| Trend | Biological Process | Expected Pattern |
|:------|:-------------------|:-----------------|
| **Linear (.L)** | Enhancer maturation, silencing progression | Monotonic Width increase or decrease |
| **Quadratic (.Q)** | Pioneer factor wave, transient stress response | Peak-then-collapse or dip-then-rise |
| **Cubic (.C)** | Circadian rhythm, cell-cycle oscillation | Two inflection points |
| **Null** | Constitutively active or stably silent | Flat over time |

## Scenario D: Tumour Heterogeneity — Single-Sample Outlier Analysis {.tabset}

**Biological context.** A cohort of 20 tumour samples with no matched
normals. You want to identify regulatory regions where individual tumours
exhibit extreme conformational departures from the cohort norm — these
may represent private driver events, subclonal regulatory rewiring, or
technical artefacts requiring QC exclusion.

### Analysis Strategy

1. Build portrait matrix on all 20 samples
2. `detect_conformational_outliers()` with robust MCD estimation on the
   full cohort → identifies domains pervasively aberrant in many samples
3. Leave-one-out analysis: for each tumour, set it as the query and the
   remaining 19 as reference — detects sample-specific outliers
4. Manhattan plot of outlier scores coloured by recurrence across samples
5. Samples driving the most outliers are flagged for QC or deeper
   characterisation

```{r scenario-d, eval = FALSE}
# Cohort-level outlier scan
cohort_outliers <- detect_conformational_outliers(
  se, method = "robust"
)

# Leave-one-out per-sample analysis
sample_outlier_counts <- sapply(colnames(se), function(s) {
  res <- detect_conformational_outliers(
    se,
    method      = "robust",
    ref_samples = setdiff(colnames(se), s)
  )
  sum(res$Outlier_Flag, na.rm = TRUE)
})

# Flag samples with excessive outliers (> 3 × median)
threshold <- 3 * median(sample_outlier_counts)
flag_samples <- names(sample_outlier_counts)[sample_outlier_counts > threshold]
cat(sprintf("%d samples flagged with > %d outliers each.\n",
            length(flag_samples), threshold))
```

---

# Best Practices & Parameter Tuning {#best-practices}

## Choosing the Right bg_quantile

The `bg_quantile` parameter controls the **Valley Trap** — the fraction of
lowest-intensity positions within each domain excluded from skewness
estimation. This prevents background noise from dominating the skewness
signal.

| Data Type | Recommended bg_quantile | Rationale |
|:----------|:------------------------|:----------|
| ChIP-seq (H3K27ac, H3K4me3) | 0.10 (default) | Moderate signal-to-noise; 10% gate removes most background |
| ChIP-seq (H3K9me3, broad marks) | 0.05 | Broad domains; too aggressive gating removes real signal tail |
| ATAC-seq | 0.10–0.15 | Nucleosome-free regions have sharp peaks; background is low |
| CUT&Tag | 0.05–0.10 | Very high signal-to-noise; lower gate preserves dynamic range |
| Low-coverage data (< 10M reads) | 0.15–0.20 | Higher gate compensates for sparser coverage |

## Normalization Method Selection

```{r norm-best-practices, eval = FALSE}
# Start with the default
se <- normalize_portrait(se, method = "TotalSignal")

# Check if batch dominates biological signal
pca_before <- plot_portrait_pca(se, feature = "Intensity", group_var = "Condition")

# If batch effects are visible, escalate to Quantile
se <- normalize_portrait(se, method = "Quantile")

# Z-score is for clustering ONLY — never feed Z-scored data to MANOVA or limma
# The Z-score transformation destroys the covariance structure that MANOVA depends on
```

**Decision flow:**

1. Try `TotalSignal` → check PCA
2. If PC1 separates by batch, not biology → try `Quantile`
3. If `Quantile` over-corrects (PC1 now separates nothing) → `TotalSignal`
   with explicit batch covariate in `test_global_shape_shift(formula = ~ batch + condition)`
4. Never use `Z-score` before differential testing

## MANOVA Test Statistic Selection

| Statistic | Robustness | Power | When to Use |
|:----------|:-----------|:------|:------------|
| **Pillai** (default) | Highest | Moderate | Most scenarios; unequal group sizes; violated assumptions |
| **Wilks** | Moderate | Highest | Balanced designs with equal covariance |
| **Hotelling-Lawley** | Low | High | Large n, balanced, few dimensions |
| **Roy** | Lowest | Special case | When you expect the effect concentrated in one dimension |

**Practical rule**: Always start with Pillai. If results look underpowered
and your design is balanced with n ≥ 8 per group, try Wilks. Rarely need
the others.

## Sample Size Guidelines

Based on `simulate_power()`:

| n per Group | Detectable Effect (eta²) | Practical Meaning |
|:------------|:-------------------------|:------------------|
| 3 | ≥ 0.30 | Only very large, overt shape shifts |
| 5 | ≥ 0.14 | Large conformational changes |
| 8 | ≥ 0.06 | Moderate shifts (typical for perturbation experiments) |
| 12+ | ≥ 0.03 | Subtle shifts; useful for patient cohort studies |

**For grant writing**: Run `simulate_power()` with your expected effect
size, then cite the resulting power curve to justify your requested
replicate number.

## Interpreting partial_eta_sq in Context

`partial_eta_sq` is the proportion of multivariate variance in the 4D
vector explained by group membership. Unlike a P-value, it does not
depend on sample size and is directly comparable across peaks.

- **0.01–0.06**: Subtle conformational shift. Biologically meaningful
  only if reproducible across independent cohorts or supported by
  orthogonal evidence (Hi-C, 4C, single-locus imaging).
- **0.06–0.14**: Moderate shift. Typically the sweet spot for
  perturbation experiments — large enough to be visually apparent in
  overlay tracks.
- **0.14–0.30**: Large shift. The peak is visibly different between
  conditions. Suitable for focused follow-up (ChIP-qPCR, reporter assays).
- **> 0.30**: Massive remodelling. Usually only seen with compound
  treatments, genetic knockouts of chromatin modifiers, or comparing
  entirely different cell types.

---

# Worked Example: Recovering Known Shape Shifts {#worked-example}

The built-in `example_se` dataset contains 500 peaks across 6 samples
(3 Control, 3 Treatment) with **5 types of ground-truth shape shifts**
embedded by simulation. This section demonstrates epiPortrait's ability
to recover these known shifts.

```{r worked-setup}
data(example_se)
se <- example_se
se <- normalize_portrait(se, method = "TotalSignal")

# Ground truth labels
true_classes <- mcols(rowRanges(se))$True_Class
cat("Ground truth composition:\n")
print(table(true_classes))
```

## Step 1: Run the Full Pipeline

```{r worked-pipeline}
shift_res <- test_global_shape_shift(se, group_var = "Condition", workers = 1)
shift_res <- classify_shape_shift(se, shift_res,
  target_group = "Treatment", ref_group = "Control")
```

## Step 2: Compare Predicted vs True Classes

```{r worked-compare}
# Merge predictions with ground truth
shift_res$True_Class <- true_classes[shift_res$Peak_ID]

# Confusion: which predicted classes correspond to which true shifts?
confusion <- table(
  Predicted = shift_res$Shape_Class[shift_res$adj.P.Val < 0.05],
  Truth     = shift_res$True_Class[shift_res$adj.P.Val < 0.05]
)
print(confusion)
```

## Step 3: Per-Class Recovery Rate

```{r worked-recovery}
# For each true class, what proportion is recovered as significant?
for (tc in unique(true_classes)) {
  idx <- which(shift_res$True_Class == tc)
  n_sig <- sum(shift_res$adj.P.Val[idx] < 0.05, na.rm = TRUE)
  cat(sprintf(
    "%-20s: %3d / %3d recovered (%.1f%%)\n",
    tc, n_sig, length(idx), 100 * n_sig / length(idx)
  ))
}
```

## Step 4: Visual Validation of the Top Hit

```{r worked-visual, fig.width = 10, fig.height = 5, eval = FALSE}
# Requires actual BigWig files on disk (example_se uses simulated paths).
top_peak <- shift_res$Peak_ID[which.min(shift_res$adj.P.Val)]
cat(sprintf("Top hit: %s (adj.P.Val = %.2e, true class = %s)\n",
            top_peak, min(shift_res$adj.P.Val), shift_res$True_Class[which.min(shift_res$adj.P.Val)]))

plot_shift_comparison(se,
  peak_id  = top_peak,
  group_var = "Condition",
  group1   = "Control",
  group2   = "Treatment",
  extend_bp = 1000
)
```

## Step 5: 4D Volcano — Where Do the Known Shifts Land?

```{r worked-volcano, fig.width = 10, fig.height = 8}
diff_res <- test_differential_feature(se,
  feature = "Intensity",
  group_var = "Condition",
  target_group = "Treatment", ref_group = "Control",
  method = "limma"
)

plot_portrait_volcano(diff_res, se,
  target_group = "Treatment",
  ref_group    = "Control",
  p_cutoff     = 0.05,
  fc_cutoff    = 1
)
```

**Key takeaway from the worked example**: Concentration and Flattening
events — where Intensity is stable but Height changes — are the scenarios
where epiPortrait provides the greatest advantage over standard
differential binding. These peaks would be invisible to DiffBind or DESeq2,
but have near-perfect recovery rates by MANOVA.

---

# Troubleshooting {#troubleshooting}

## MANOVA Returns All P-values = 1 or All P-values = 0

| Symptom | Cause | Solution |
|:--------|:------|:---------|
| All adj.P.Val = 1 | Zero within-group variance in ≥ 3 of 4 dimensions (e.g., identical samples) | Check input data; ensure BigWig files are not identical copies |
| All adj.P.Val = 1 | Wrong `term` specified for formula | Use the auto-selected term or explicitly name the correct model term |
| All P.Val < 1e-10 | Groups are completely separated (e.g., different cell types) | This is expected; use `partial_eta_sq` to rank effect sizes instead |
| `"MANOVA_Error"` in Status column | Singular covariance matrix for that peak | That peak has near-zero signal in all samples; these are automatically flagged and do not affect other peaks |

## Skewness Values Are All Near Zero

- Check that `bg_quantile` is not too high (e.g., > 0.3). If the Valley
  Trap removes too much signal, the remaining positions may all have
  similar values, producing skewness ≈ 0.
- Verify that your peaks contain enough positions. For very narrow peaks
  (e.g., < 50 bp), the number of base-pair-resolution BigWig positions
  may be insufficient for robust skewness estimation. Consider using
  `stitch_epi_peaks()` to create macro-domains first.

## Memory Exhaustion with Large BigWig Files

- Set `on_disk = TRUE` to use HDF5-backed matrices. This moves all assay
  data out of RAM.
- Set `cache_dir` to a persistent directory. The first run is
  memory-intensive (BigWig import), but subsequent runs with the same
  peaks skip I/O entirely.
- Reduce `workers` to 1 for memory-constrained environments. Multiple
  workers each load their own BigWig data into memory.

## MCD Covariance Estimation Fails

```
Error: Only 7 valid domains; need at least 8 for reliable
covariance estimation.
```

- Switch to `method = "classic"` when the dataset is too small for MCD.
  The classic estimator only needs n > 4 (the number of dimensions).
- Alternatively, increase `collapse_fun` aggregation by providing more
  reference samples.

## Classification Returns Mostly "Stable" Despite Significant MANOVA

- Lower `fc_threshold` from the default 0.5 to 0.3. MANOVA may detect
  a consistent but subtle shift across multiple dimensions that none
  individually exceeds 0.5 log2FC.
- Check that your `target_group` and `ref_group` match the levels in
  `colData(se)`.
- If `Skewness_Delta` is consistently near zero, your data may have no
  directional asymmetry — this is common in symmetric peaks and is not
  a bug.

## Paired Design Gives Fewer Results Than Unpaired

MANOVA with a blocking factor consumes additional degrees of freedom
(1 per block after the first). With n = 3 pairs, the effective residual
df for MANOVA is 1, which reduces power substantially. Consider:
- Using `test_differential_feature()` with `method = "limma"` and
  `block_var` for per-dimension tests in small paired designs
- Increasing sample size if paired MANOVA is essential

---

# Reproducibility

All analyses in this vignette were conducted with the following software
versions. For reproducible results, set a seed (`set.seed()`) before running
simulation-based functions (`simulate_power()`, `detect_conformational_outliers()`
with MCD, etc.).

```{r session-info}
sessionInfo()
```
